Leadership across cultures and contexts: an exploratory analysis

### 2020-06-23 leadership_across_cultures_contexts.Rmd
More reductions.

Despite decades of research and theorizing on leadership, to our knowledge there is no systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most scholarship on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

There is potentially more diversity within "non-WEIRD" populations than within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. Comparative research must assess cultural variation without assuming that all members within a population are essentially similar, and without putting undue emphasis on either the between-population differences or similarities.

@@

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. One outstanding question is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies have focused on community or political leaders, often overlooking leadership across contexts and levels of social organization, such as among kin and social groups. A second outstanding question is what benefits *and* costs both leaders and followers incur. While the benefits leaders accrue have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated [but see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, community leadership is typically informal and emerges in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015]. Leaders also play important roles in ritual and religeous contexts [@singh_cultural_2017; @winkelman_crosscultural_2020], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Leadership in between-group conflict and cooperation is also common across nonindustrial societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community level, and institutionalized managerial roles are associated with sedentarization, defensibility of resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of resources creates increased opportunities for resource management and is often associated with increased territoriality and inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

We first examined the leadership dimensions with the most evidence, i.e., the  `r commondim_n` for which evidence was found in at least 60% of cultures. Of these, `r universals_n` dimensions did not meaningfully vary by continental region, subsistence strategy, group context, or leader sex, and are therefore candidates for universal dimensions of human leadership. See Table \@ref(tab:leaderuniversals).

@@

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over the intercept-only models (Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying leadership dimensions, one factor had a noticeably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions (Figure \@ref(fig:heatmapdrop1)).

@@

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, there were three leadership dimensions in which subsistence strategy played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For group context, region, and leader sex see Figures S\@ref(fig:heatmapgroups), S\@ref(fig:heatmapregion), and S\@ref(fig:heatmapsex).

@@

*High status* was the most frequently identified leader quality in the sample (`r statusPercTxt`% of text records and `r statusPercCult`% of cultures provided evidence of high status leaders). To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status leaders. These included *`r paste(rev(highstatus_plot$data$Word)[1:5], collapse=', ')`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were the strongest predictive terms. Notable weak-predictor terms included, *influence*, *authority*, *headman*, and *chief*, and *economic* was a negative predictor. See Figure \@ref(fig:highstatusplot)A.

@@

This exploratory study systematically evaluated ethnographic evidence for `r length(all_study_vars)` leadership dimensions from a sample of `r nrow(leader_cult)` largely nonindustrial cultures including hunter-gatherers, pastoralists, horticulturalists, and agriculturalists. Results revealed universal dimensions of leadership as well as important variation by group context, subsistence type, continental region, and leader sex.

@@

Some common leadership dimensions were also relatively invariant across group context, subsistence strategy, continental region, and leader sex. These dimensions included the qualities, knowledgeable or intelligent, wealthy, strategically nepotistic, and generous; and functions, resolving conflicts and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal" we do not mean all leaders have all these traits. Instead, we mean there is evidence for these traits across a solid majority of cultures that do not vary much by key measures of population or group variation (i.e., the four factors) (Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport)).

@@

These potentially universal dimensions of leadership correspond to several evolutionary theories that have drawn on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature corresponds, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2020].

>>>>

Despite decades of scholarship on leadership, to our knowledge there is no systematic investigation of leader functions and qualities, and the costs and benefits of leadership and followership, across a representative sample of nonindustrial societies. Moreover, outside of anthropology, most scholarship on leadership is based on data from Western or Westernized, postindustrial societies [@von_rueden_leadership_2015].

@@

There is potentially more diversity within "non-WEIRD" populations than WEIRD ones, and there is also substantial overlap of the behaviors all people [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of societies. Comparative research must assess cultural variation without assuming that all members within a population are essentially similar, and without putting undue emphasis on either the between-population differences or similarities.

@@

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over group decision-making [@von_rueden_leadership_2015], whatever the group may be. One outstanding question is the degree to which theoretically important dimensions of leadership vary between cultures and across social contexts. Leadership studies have focused on community and political leaders, often overlooking leadership across contexts and levels of social organization, such as among kin and social groups. A second outstanding question is what benefits *and* costs leaders and followers incur. While the benefits leaders accrue have received substantial attention -- fitness benefits in particular -- the costs of leadership and followership have been relatively under-investigated [but see @cheng_dominance_2019; @price_evolution_2014; @glowacki_leadership_2015; @von_rueden_leadership_2014].

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, community leadership is typically informal and emerges in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015]. Leaders also play important roles in ritual and religeous contexts [@singh_cultural_2017; @winkelman_crosscultural_2020], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Leadership in between-group conflict and cooperation is also common across nonindustrial societies, including hunter-gatherers [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @apicella_social_2012].

@@

<!-- MANAGEMENT FUNCTION IN MORE CULTUARLL COMPLEX -->
Stronger authority of community leaders, leadership beyond the community, and institutionalized managerial roles are associated with sedentarization, defendable resources, and social stratification [@johnson_evolution_1987; @johnson_organizational_1982; @kaplan_evolutionary_2009; @ames_hierarchies_1985]. Among horticulturalists and agriculturalists, managerial leadership can promote more efficient use of shared resources such as water reserves and cooperative labor [e.g., @von_rueden_leadership_2014]. Pastoralists and coastal populations often face similar pressures concerning grazing lands and fishing access [e.g., @widmer_evolution_1988; @stevens_sherpa_1990]. 
<!-- In the wake of economic intensification, the functions of leadership becomes increasingly associated with managing economic systems and distribution of resources.  -->
Defensibility of resources creates increased opportunities for resource management and is often associated with increased territoriality and inter-group conflicts [@glowacki_evolutionary_2017]. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military operations [@lopez_evolutionary_2017].

@@

We first examined the leadership dimensions with the most evidence, i.e., the  `r commondim_n` for which evidence was found in at least 60% of cultures. Of these, `r universals_n` dimensions did not meaningfully vary by continental region, subsistence strategy, group context, or leader sex, and are therefore candidates for universal human leadership dimensions. See Table \@ref(tab:leaderuniversals).

@@

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over intercept-only models (Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which factor contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying leadership dimensions, one factor had a noticeably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions (Figure \@ref(fig:heatmapdrop1)).

@@

For each of our four factors, we then investigated which levels of the factor were associated with higher or lower levels of evidence for the leadership dimension. For instance, there were three dimensions in which subsistence strategy played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to other subsistence strategies (and averaging over levels of the other three factors), and pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For group context, region, and leader sex see Figures S\@ref(fig:heatmapgroups), S\@ref(fig:heatmapregion), and S\@ref(fig:heatmapsex).

@@

*High status* was the most frequently identified leader quality (`r statusPercTxt`% of text records and `r statusPercCult`% of cultures provided evidence of high status leaders). To explore factors that distinguished these records from those with no evidence for high status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with the lasso penalty, $\alpha = 1$) of *High status* (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provided evidence of high status leaders. These included *`r paste(rev(highstatus_plot$data$Word)[1:5], collapse=', ')`* and *`r rev(highstatus_plot$data$Word)[6]`*, which were the strongest predictive terms. Notable weak-predictor terms included, *influence*, *authority*, *headman*, and *chief*, and *economic* was a negative predictor. See Figure \@ref(fig:highstatusplot)A.

@@

This exploratory study systematically evaluated ethnographic evidence for `r length(all_study_vars)` leadership dimensions from a diverse sample of `r nrow(leader_cult)` largely nonindustrial cultures. Results revealed universal dimensions of leadership as well as important variation by group context, subsistence type, continental region, and leader sex.

@@

Some common leadership dimensions were also relatively invariant across group context, subsistence strategy, continental region, and leader sex. These dimensions included the qualities, knowledgeable/intelligent, wealthy, strategically nepotistic, and generous; and functions, resolving conflicts and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal" we do not mean all leaders have all these traits. Instead, we mean there is evidence for these traits across a solid majority of cultures that do not vary much by key measures of population or group variation (i.e., the four factors) (Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport)).

@@

These potentially universal dimensions of leadership correspond to several evolutionary theories that have drawn on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature corresponds in part to leaders who serve as models for their followers [see also @garfield_investigating_2020].
